---
title: | 
  | Accessing and Analyzing 
  | MLB Pitch Tracking Data in R
# subtitle: "Subtitle"
author: "Aaron R. Baggett, Ph.D."
date: "March 28, 2017"
institute: | 
  | University of Mary Hardin-Baylor
  | Department of Psychology
output: 
  beamer_presentation: 
    highlight: tango
    includes:
      in_header: header.tex
    keep_tex: yes
    pandoc_args: --latex-engine=xelatex
    # theme: Antibes
monofont: Lucida Console # Comment out for default
mainfont: Roboto # Comment out for default
fontsize: 14pt
classoption: aspectratio = 1610
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
load("~/Dropbox/Research Projects/Austin_R_Meetup/data/atbat.Rda")
load("~/Dropbox/Research Projects/Austin_R_Meetup/data/pitch.Rda")
load("~/Dropbox/Research Projects/Austin_R_Meetup/data/umpire.Rda")
```

## Resources
- Slides, data, and R code are available at:

\begin{center}
\color{brightpink}\Large{\texttt{bit.ly/austin\_r}}
\end{center}

# Major League Baseball (MLB) \protect \newline Pitch Tracking Data

## MLB Pitch Tracking Data
- Since 2007, MLB has tracked pitch location and play-by-play data for all games
- Source: Sportvision PITCHf/x system
- PITCHf/x data are fed real time to mobile and desktop apps
- All data are stored in XML format on MLB servers

## MLB Pitch Tracking Data
- Location: \color{brightpink}\url{http://gd2.mlb.com/components/game/mlb/}

\vspace{-1mm}
\begin{center}
\includegraphics[width = 4.5in]{../figs/gd_inning_all.png}
\end{center}

# Accessing MLB Pitch Tracking Data

## Accessing MLB Pitch Tracking Data
- R packages:
    1. *`pitchRx`*: Data collection
    2. *`dplyr`*: Data analysis

## `pitchRx`
- Prior to 2013, researchers had to scrape PITCHf/x data manually
- In 2013, Carson Sievert created the *`pitchRx`* R package
- *`pitchRx`* contains tools for accessing play-by-play data

## `pitchRx`
- Prior to 2013, researchers had to scrape PITCHf/x data manually
- In 2013, Carson Sievert created the *`pitchRx`* R package
- *`pitchRx`* contains tools for accessing play-by-play data

\footnotesize
```{r, eval = FALSE}
pfx_db <- src_sqlite("pfx_db.sqlite3", create = TRUE)
files <- c("inning/inning_all.xml", "players.xml", "miniscoreboard.xml")
scrape(start = "YYY-MM-DD", end = "YYY-MM-DD", suffix = files, 
  connect = pfx_db$con)
pfx_db <- src_sqlite("~/your/working/directory/pfx_db.sqlite3")
src_tbls(pfx_16)
```

## `pitchRx`
- Once we set up a PITCHf/x database, we have access to all MLB pitch and gameplay data
- Best to use a small date range for initial setup
- 10 primary tables in the data

## PITCHf/x Data Tables
\small
\begin{table}[ht]
\centering
\begin{tabular}{ll}
\hline
Table Name & Description \\ 
\hline
{\texttt{action}} & Ball/strike count, result of pitch, ... \\ 
{\texttt{atbat}} & Pitcher/batter names, handedness, heights, at bat result, ... \\ 
{\texttt{coach}} & Names of manager and staff, ... \\ 
{\texttt{game}} & Venue, start time, time zone, TV, win-loss records, ... \\ 
{\texttt{media}} & Mobile/TV media assets, ... \\ 
{\texttt{pitch}} & Umpire's decision/outcome, strike zone parameters, x-y coordinates, ... \\ 
{\texttt{player}} & Players' stats, position, number, ... \\ 
{\texttt{po}} & Details about put out attempts (e.g., pickoffs and stolen bases), ... \\ 
{\texttt{runner}} & Details about base runner(s) and at bat events, ... \\ 
{\texttt{umpire}} & Umpire names and positions, ... \\ 
\hline
\end{tabular}
\end{table}

## PITCHf/x Data Tables
For most analyses, we usually work with:

\small
\begin{table}[ht]
\centering
\begin{tabular}{ll}
\hline
Table Name & Description \\ 
\hline
{\texttt{action}} & Ball/strike count, result of pitch, ... \\ 
\rowcolor{SpringGreen}{\texttt{atbat}} & Pitcher/batter names, handedness, heights, at bat result, ... \\ 
{\texttt{coach}} & Names of manager and staff, ... \\ 
{\texttt{game}} & Venue, start time, time zone, TV, win-loss records, ... \\ 
{\texttt{media}} & Mobile/TV media assets, ... \\ 
\rowcolor{SpringGreen}{\texttt{pitch}} & Umpire's decision/outcome, strike zone parameters, x-y coordinates, ... \\ 
{\texttt{player}} & Players' stats, position, number, ... \\ 
{\texttt{po}} & Details about put out attempts (e.g., pickoffs and stolen bases), ... \\ 
{\texttt{runner}} & Details about base runner(s) and at bat events, ... \\ 
\rowcolor{SpringGreen}{\texttt{umpire}} & Umpire names and positions, ... \\ 
\hline
\end{tabular}
\end{table}

## `dplyr`
- Wickham and Francois (2016)
- A grammar of data manipulation
- Provides a set of verbs for lots of tasks
    - *`select()`*: Selects columns
    - *`filter()`*: Filters rows (e.g., `==`, `!=`, `<=`, etc.)
    - *`arrange()`*: Re-orders and sorts rows
    - *`mutate()`*: Creates new variables/columns
    - *`summarise()`*: Summarizes values/output
    - *`group_by()`*: Allows for by-group operations

## Accessing MLB Pitch Tracking Data
- R packages:
    1. *`pitchRx`*: Data collection
    2. *`dplyr`*: Data analysis

# Analyzing MLB Pitch Tracking Data

## Tonight
- There are several ways to analyze PITCHf/x data
    - Ex.: Pitching/batting outcomes, predictive models, et al.
- Tonight though, let's concentrate on home plate umpire decisions
- Specifically:
    1. How many pitches do umpires see during games? Of those, how many require a decision?
    2. How accurate are *all* umpires over the season? How accurate are *individual* umpires over the season?

## 1. Pitches Seen vs. Decisions Made
- How many pitches do umpires see during games? Of those, how many require an umpire decision?
    - **Pitches seen:** Total number of recorded pitches thrown during game
    - **Decisions made:** Total number of called strikes and called balls during game

## 1. Pitches Seen vs. Decisions Made
- We'll use the `pitch` table to answer these questions
- Steps:
    1. Create data frame for pitches seen, *`observed`*
    2. Create data frame for decisions made, *`decisions`*
    3. Join *`observed`* and *`decisions`*
    4. Calculate proportion of pitches requiring decision
    5. Calculate simple descriptive statistics

## 1. Pitches Seen vs. Decisions Made
- Step 1. Create data frame for pitches seen

\footnotesize
```{r, eval = FALSE}
observed <- pitch %>% 
  group_by(gameday_link) %>% 
  summarize(seen = n())
```

<!--
## 1. Pitches Seen vs. Decisions Made
- Step 1. Create data frame for pitches seen
    - `pitch`: Current data frame

\footnotesize
```{r, eval = FALSE}
observed <- pitch %>% 
  group_by(gameday_link) %>% 
  summarize(seen = n())
```

## 1. Pitches Seen vs. Decisions Made
- Step 1. Create data frame for pitches seen
    - `pitch`: Current data frame
    - `group_by()`, `summarize()`, `n()`: *`dplyr`* verbs

\footnotesize
```{r, eval = FALSE}
observed <- pitch %>% 
  group_by(gameday_link) %>% 
  summarize(seen = n())
```

## 1. Pitches Seen vs. Decisions Made
- Step 1. Create data frame for pitches seen
    - `pitch`: Current data frame
    - `group_by()`, `summarize()`, `n()`: *`dplyr`* verbs
    - `gameday_link`: Unique date/team label

\footnotesize
```{r, eval = FALSE}
observed <- pitch %>% 
  group_by(gameday_link) %>% 
  summarize(seen = n())
```

## 1. Pitches Seen vs. Decisions Made
- Step 1. Create data frame for pitches seen
    - `pitch`: Current data frame
    - `group_by()`, `summarize()`, `n()`: *`dplyr`* verbs
    - `gameday_link`: Unique date/team label
    - `seen`: New name for variable `n()`

\footnotesize
```{r, eval = FALSE}
observed <- pitch %>% 
  group_by(gameday_link) %>% 
  summarize(seen = n())
```
-->

## 1. Pitches Seen vs. Decisions Made
- Step 1. Create data frame for pitches seen
- R code:
    - *`pitch`*: Current data frame
    - *`group_by()`*, *`summarize()`*, *`n()`*: *`dplyr`* verbs
    - *`gameday_link`*: Unique date/team label
    - *`seen`*: New name for variable `n()`
    - *`observed`*: Name of new data frame

\footnotesize
```{r, eval = FALSE}
observed <- pitch %>% 
  group_by(gameday_link) %>% 
  summarize(seen = n())
```

## 1. Pitches Seen vs. Decisions Made
- Step 1. Create data frame for pitches seen

\footnotesize
```{r}
(observed <- pitch %>% 
  group_by(gameday_link) %>% 
  summarize(seen = n()))
```

## 1. Pitches Seen vs. Decisions Made
- Step 2. Create data frame for decisions made
- We need to omit all pitches/outcomes except for called strikes and called balls

\footnotesize
```{r, eval = FALSE}
decisions <- pitch %>% 
  group_by(gameday_link) %>% 
  filter(des == "Called Strike" | des == "Ball") %>% 
  summarize(decisions = n())
```

## 1. Pitches Seen vs. Decisions Made
- Step 2. Create data frame for decisions made
- We need to omit all pitches/outcomes except for called strikes and called balls
- R code:
    - *`filter()`*: Returns rows with matching conditions

\footnotesize
```{r, eval = FALSE}
decisions <- pitch %>% 
  group_by(gameday_link) %>% 
  filter(des == "Called Strike" | des == "Ball") %>% 
  summarize(decisions = n())
```

## 1. Pitches Seen vs. Decisions Made
- Step 2. Create data frame for decisions made

\footnotesize
```{r}
(decisions <- pitch %>% 
  group_by(gameday_link) %>% 
  filter(des == "Called Strike" | des == "Ball") %>% 
  summarize(decisions = n()))
```

## 1. Pitches Seen vs. Decisions Made
- Step 3. Join *`observed`* and *`decisions`* by *`gameday_link`*

\footnotesize
```{r, eval = FALSE}
pitches <- inner_join(observed, decisions, by = "gameday_link")
```

## 1. Pitches Seen vs. Decisions Made
- Step 3. Join *`observed`* and *`decisions`* by *`gameday_link`*
- R code:
    - *`inner_join()`*: Returns observations that match in both x and y

\footnotesize
```{r, eval = FALSE}
pitches <- inner_join(observed, decisions, by = "gameday_link")
```

## 1. Pitches Seen vs. Decisions Made
- Step 3. Join *`observed`* and *`decisions`* by *`gameday_link`*

\footnotesize
```{r}
(pitches <- inner_join(observed, decisions, by = "gameday_link"))
```

## 1. Pitches Seen vs. Decisions Made
- Step 4. Calculate proportion of pitches requiring decision

\footnotesize
```{r, eval = FALSE}
pitches <- pitches %>% 
  mutate(prop = decisions/seen)
```

## 1. Pitches Seen vs. Decisions Made
- Step 4. Calculate proportion of pitches requiring decision
- R code: 
    - *`mutate()`*: Adds new variable

\footnotesize
```{r, eval = FALSE}
pitches <- pitches %>% 
  mutate(prop = decisions/seen)
```

## 1. Pitches Seen vs. Decisions Made
- Step 4. Calculate proportion of pitches requiring decision

\footnotesize
```{r}
(pitches <- pitches %>% 
  mutate(prop = decisions/seen))
```

## 1. Pitches Seen vs. Decisions Made
- Step 5. Calculate simple descriptive statistics

\footnotesize
```{r}
(pitch_summs <- pitches %>% 
  summarize(m_pitches = mean(seen),
    sd_pitches = sd(seen),
    m_calls = mean(decisions),
    sd_calls = sd(decisions),
    m_prop = mean(prop),
    sd_prop = sd(prop)))
```

## Umpire Accuracy
- Check out `umpire_accuracy.R` in my GitHub Repo for this talk
- Overall, umpires are quite accurate

\begin{table}[ht]
\centering
\begin{tabular}{lllr}
\hline
\em{M} & \em{SD} & \em{SEM} & 95\% CI \\
\hline
0.94 & 0.24 & 0.05 & [0.84, 0.98] \\
\hline
\end{tabular}
\end{table}

## Umpire Accuracy
- Here's a plot of the cumulative accuracy for MLB umpires over the season

\vspace{-1mm}
\begin{center}
\includegraphics[width = 5in]{../figs/ump_acc_season.pdf}
\end{center}

# Questions?

## Contact Details
\begin{center}
{\Large{\bf{Aaron R. Baggett, Ph.D.}}}

{\small{Assistant Professor of Psychology\\
University of Mary Hardin-Baylor\\
{\large{\faEnvelope}} \url{abaggett@umhb.edu}\\
{\large{\faPhone}} (254) 295-4553\\
{\large{\faTwitter}} \href{http://twitter.com/aaron_baggett}{@aaron\_baggett}}}
\end{center}

